---
title: "R2 -- human"
author: "RA ([numpde](https://github.com/numpde/))"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
editor_options:
 chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Preliminaries {.tabset .tabset-fade}


### Introduction

--
  

### Setup

```{r}
suppressPackageStartupMessages({
  # install.packages("kableExtra")
  requireNamespace("kableExtra")

  # install.packages("crunch")
  requireNamespace("crunch")

  # https://rstudio.github.io/DT/
  # install.packages("DT")
  requireNamespace("DT")

  # may need to say "don't restart session"
  # install.packages("RCurl")
  requireNamespace("RCurl")

  # install.packages("dplyr")
  library(dplyr)
  # install.packages("ggplot2")
  library(ggplot2)

  #
  requireNamespace("tidyr")

  # BiocManager::install("org.Mm.eg.db")
  requireNamespace("org.Mm.eg.db")

  # https://www.bioconductor.org/packages/release/bioc/vignettes/goseq/inst/doc/goseq.pdf
  # BiocManager::install("goseq")
  requireNamespace("goseq")

  # https://vroom.r-lib.org/articles/vroom.html
  # install.packages("vroom")
  requireNamespace("vroom")

  # BiocManager::install("DESeq2")
  requireNamespace("DESeq2")

  # BiocManager::install("scater")
  requireNamespace("scater")

  # BiocManager::install("scran")
  requireNamespace("scran")

  # BiocManager::install("igraph")
  requireNamespace("igraph")

  # install.packages("pheatmap")
  requireNamespace("pheatmap")
  # install.packages("stringr")
  requireNamespace("stringr")
  # install.packages("reshape2")
  requireNamespace("reshape2")
  # install.packages("pathlibr")
  requireNamespace("pathlibr")
  # install.packages("pbapply")
  requireNamespace("pbapply")

  # install.packages("Rtsne")
  requireNamespace("Rtsne")
  # install.packages("uwot")
  requireNamespace("uwot")

  # avoid importing `exprs` that leads to clashes
  requireNamespace("rlang")
})
```

### SessionInfo

```{r}
sessionInfo()
```

### Paths

```{r}
BASEPATH <- pathlibr::Path$new(".")
stopifnot("main.Rmd" %in% names(BASEPATH$dir))
```

```{r}
path_to <- (function(.) Sys.glob(BASEPATH$join("../..")$join(.)$show))
```

### Plotting

```{r}
ggplot2::theme_set(theme_light(base_size = 15))
```

```{r}
kable <- function(.) {
  kableExtra::kbl(., align = "c") %>%
    kableExtra::kable_paper("hover", full_width = F)
}
```

```{r}
options(DT.options = list(
  pageLength = 16,
  language = list(search = "Filter:")
))
```

```{r}
link <-
  function(x) {
    dplyr::case_when(
      # If x is like GO:003453
      startsWith(x, "GO:") ~ paste0(
        "<a target='_blank'",
        "href='", "http://amigo.geneontology.org/amigo/term/", x, "'",
        ">", x, "</a>"
      ),
      # Otherwise try this
      TRUE ~ paste0(
        "<a target='_blank'",
        "href='", "http://www.informatics.jax.org/marker/summary?nomen=", x, "'",
        ">", x, "</a>"
      )
    )
  }
```

### Dataframes

```{r}
# Use first column as index
by_col1 <- (function(.) tibble::column_to_rownames(., colnames(.)[1]))
# Use index as new column `name`
ind2col <- (function(., name) tibble::rownames_to_column(., var = name))
```

```{r}
arrange.like <- (function(X, T) X[rownames(T), , drop = FALSE])
rows.of <- (function(rows, X) X[rows, , drop = FALSE])
```

```{r}
Sys.setenv("VROOM_CONNECTION_SIZE" = 131072 * 4)
from_csv <- function(.) {
  vroom::vroom(., del = "\t", progress = FALSE) %>%
    by_col1() %>%
    suppressMessages()
}
```


### Misc

```{r}
# Wrap functions that do `library(...)`
# and force-detach all attached packages
unattach_packages <- function(f) {
  # install.packages("crunch")
  requireNamespace("crunch")

  # wrapper for f
  function(...) {
    base::with(
      crunch::ContextManager(
        enter = base::search,
        exit = function() {
          for (pkg in setdiff(base::search(), enter)) {
            base::detach(pkg, character.only = TRUE, force = TRUE)
          }
        },
        # this overwrites the variable `enter`, if existent
        as = "enter"
      ),
      {
        f(...)
      }
    )
  }
}

# Usage:
# unattach_packages(function(x, y) { library(goseq); x + y })(1, 2)
```




## Datasets {.tabset .tabset-fade .tabset-pills}

### Notes

This is just a backup of 
the human part of "R2".

### AB -- Human M1 (10x) {.tabset .tabset-fade}

#### Load

Single-nucleus transcriptomes from 76533 nuclei
[[src](https://portal.brain-map.org/atlases-and-data/rnaseq/human-m1-10x)].

```{r, cache=TRUE}
# AllenBrain-Human M1
sce_abh_m1 <- SingleCellExperiment::SingleCellExperiment(
  assays = list(
    counts = (
      path_to("*/AllenBrain/Human-M1*/f*cells/data.*.gz") %>%
        from_csv() %>%
        as.matrix()
    )
  ),
  colData = (
    path_to("*/AllenBrain/Human-M1*/f*cells/meta.*.gz") %>%
      from_csv()
  )
)
```

#### Cell types

```{r}
exclude_celltypes <- c("Astro", "Endo", "Micro-PVM", "OPC")
```

```{r}
sce_abh_m1$subclass_label %>%
  data.frame(x = .) %>%
  ggplot() +
  aes(x = x, fill = if_else(x %in% exclude_celltypes, "drop", "keep")) +
  geom_bar() +
  scale_y_log10() +
  # hist_theme +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_text(a = 0, hjust = 0.5),
    legend.title = element_blank(),
  )
```

Filter cell types.

```{r, cache=TRUE}
sce_abh_m1 <-
  sce_abh_m1 %>% {
    (.)[, !((.)$subclass_label %in% exclude_celltypes)]
  }
```

#### Diagnostics


```{r, cache=TRUE}
sce_abh_m1 <-
  sce_abh_m1 %>%
  {
    scater::addPerCellQC(
      x = (.),
      subsets = list(Mito = grep("mt-", rownames(.)))
    )
  } %>%
  scater::logNormCounts() %>%
  scater::runPCA(name = "PCA", ncomponents = 25) %>%
  scater::runTSNE(perplexity = 10, dimred = "PCA") %>%
  scater::runUMAP()
```


```{r}
sce_abh_m1 %>%
  {
    function(dimred) {
      scater::plotReducedDim(object = (.), dimred = dimred)
    }
  } %>%
  {
    gridExtra::grid.arrange((.)("PCA"), (.)("UMAP"), nrow = 1)
  }
```




```{r include=FALSE}
knitr::knit_exit()
```








## Notes

"echo = FALSE": prevent printing of the R code that generated the plot.

