---
title: "Apoe in Tam & al."
author: "RA ([numpde](https://github.com/numpde/))"
date: "`r format(Sys.time(), '%Y-%m-%d')`"
output: 
  html_document:
    toc: true
    toc_depth: 2
editor_options:
 chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction


[Tam et al, 2019](https://doi.org/10.1016/j.celrep.2019.09.066)

## Preliminaries {.tabset .tabset-fade .tabset-pills}

### Setup

```{r}
suppressPackageStartupMessages({
  # install.packages("kableExtra")
  requireNamespace("kableExtra")

  # install.packages("dplyr")
  library(dplyr)
  # install.packages("ggplot2")
  library(ggplot2)

  # BiocManager::install("DESeq2")
  requireNamespace("DESeq2")

  # install.packages("pheatmap")
  requireNamespace("pheatmap")
  # install.packages("stringr")
  requireNamespace("stringr")
  # install.packages("reshape2")
  requireNamespace("reshape2")
  # install.packages("pbapply")
  requireNamespace("pbapply")
  # install.packages("pathlibr")
  requireNamespace("pathlibr")
  # install.packages("stringdist")
  requireNamespace("stringdist")

  # avoid importing `exprs` that leads to clashes
  requireNamespace("rlang")
})
```

```{r}
`%<>%` <- magrittr::`%<>%` 
```

### SessionInfo

```{r}
sessionInfo()
```


### Paths

```{r}
BASEPATH <- pathlibr::Path$new(".")
stopifnot("main.Rmd" %in% names(BASEPATH$dir))
```

```{r}

path_to <- (function(.) Sys.glob(BASEPATH$join("../..")$join(.)$show))
```

### Plotting

```{r}
kable <- function(.) {
  kableExtra::kbl(., align = "c") %>%
    kableExtra::kable_paper("hover", full_width = F)
}
```


## Datasets

### Tam et al, 2019

[ExpressionSet](https://bioconductor.riken.jp/packages/3.1/bioc/vignettes/Biobase/inst/doc/ExpressionSetIntroduction.pdf)

```{r}
eset_tam <- Biobase::ExpressionSet(
  assayData = (
    path_to("*/*Tam*/*/GSE124439_data*") %>%
      read.csv(sep = "\t", check.names = FALSE) %>%
      tibble::column_to_rownames("gene/TE") %>%
      as.matrix()
  ),
  phenoData = Biobase::AnnotatedDataFrame(
    data = (
      cbind(
        path_to("*/*Tam*/*/GSE124439_meta*") %>%
          read.csv(sep = "\t") %>%
          tibble::column_to_rownames("hra"),
        path_to("*/*Tam*/*/*/S1A_NYGC_clinical.csv*") %>%
          read.csv(sep = "\t", comment.char = "#") %>%
          tibble::column_to_rownames("hra") %>%
          # Drop duplicate columns:
          select(-sample_group, -tissue)
      )
    )
  )
)
```

```{r}
eset_tam@phenoData@data$sample_group %<>% factor
```

Sample groups are:
`r paste(levels(eset_tam@phenoData@data$sample_group), collapse = ", ")`.


## Analysis

### [DESeq2](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html)

```{r}
dds_tam <- DESeq2::DESeqDataSetFromMatrix(
  countData = eset_tam@assayData$exprs,
  colData = eset_tam@phenoData@data,
  design = ~sample_group
)
```

Only keep two conditions for DE analysis.

```{r}
dds_tam <-
  dds_tam[, dds_tam$sample_group %in% c("ALS", "non_neuro")]
```

There are now
`r dds_tam@colData$subject_id %>% unique %>% length`
samples from
`r dds_tam@colData$subject_id %>% length`
subjects.

DESeq [note](https://bioconductor.org/packages/release/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#factorlevels) on factor levels.

```{r}
dds_tam$sample_group %<>%
  relevel(ref = "non_neuro") %>%
  droplevels()
```

Attach the DESeq normalization
and
get normalized counts.

```{r}
dds_tam %<>%
  DESeq2::estimateSizeFactors()
```

```{r}
dds_tam.norm_counts <-
  dds_tam %>%
  DESeq2::counts(normalized = TRUE)
```

DESeq

```{r}
dds_tam <-
  DESeq2::DESeq(dds_tam)
```

```{r}
dds_tam@colData$ALS_subtype %<>% 
  as.factor() %>%
  addNA()

dds_tam@colData$gender %<>% 
  as.factor()
```

```{r}
de_results <-
  DESeq2::results(dds_tam, alpha = 0.05) %>%
  as.data.frame()

de_results %>%
  slice_min(padj, n = 10, with_ties = TRUE) %>%
  kable()
```

`plotCounts` shows
normalized counts plus a pseudocount of 0.5.

```{r}
dds_tam %>% 
  DESeq2::plotCounts(gene = "ANKRD20A9P", intgroup = "ALS_subtype")
```

```{r}
dds_tam %>% 
  DESeq2::plotCounts(gene = "MAPK8IP1P2", intgroup = "ALS_subtype")
```

```{r}
dds_tam %>% 
  DESeq2::plotCounts(gene = "ARAP3", intgroup = "ALS_subtype")
```

```{r}
dds_tam %>% 
  DESeq2::plotCounts(gene = "APOE", intgroup = "ALS_subtype")
```

```{r}
dds_tam %>% 
  DESeq2::plotCounts(gene = "APOE", intgroup = "gender")
```




The following are the most predictive marker genes
from each category
proposed by 
[Tam et al, 2019](https://www.sciencedirect.com/science/article/pii/S221112471931263X#mmc4),
extracted from their
[Table S2](https://ars.els-cdn.com/content/image/1-s2.0-S221112471931263X-mmc4.xlsx).

```{r}
tam_markers <- c(
  # ALS-Ox
  c("CTXN3", "LINC00507", "CALB1", "GPR22", "TMEM155", "MAL2", "PCP4"),
  # ALS-Glia
  c("NKX6-2", "SERPINA3", "KLK6", "SLC5A11", "ITGA2", "AQP1", "TNC"),
  # ALS-TE
  c("AP1G2", "CORO6", "FAM153B", "NPIPB3", "FAM153A", "SRRM3", "LINC00599")
)
```

For a suitably transformed expression table,
those genes 
indeed cluster
into the same three sets.
In particular, APOE clusters with the ALS-Glia subtype.

```{r, out.width = "100%"}
dds_tam.norm_counts[c(tam_markers, "MAPK8IP1P2", "ANKRD20A9P", "APOE"), ] %>%
  {
    log2(. + 0.5)
  } %>%
  # Standardize rows
  t() %>%
  scale() %>%
  t() %>%
  {
    (.)[, order(dds_tam@colData$ALS_subtype)]
  } %>%
  pheatmap::pheatmap(
    labels_col = "",
    cluster_cols = FALSE,
    annotation_col = (
      dds_tam@colData %>% as.data.frame() %>% select(sample_group, ALS_subtype)
    )
  )
```


***


```{r include=FALSE}
knitr::knit_exit()
```



## Notes

"echo = FALSE": prevent printing of the R code that generated the plot.
